Anthropogenic plant communities of the Iberian Atlantic ecoregion

### 2024-04-26 doc/manuscript.Rmd
readme

The total synanthropic species pool (n = 1,162) was composed of 78 % natives (n = 908), 15 % putative archaeophytes (n = 178) and 7 % neophytes (n = 76). The most frequent putative archaeophytes were *Capsella bursa-pastoris* (390 occurrences), *Senecio vulgaris* (332), *Anisantha sterilis* (301), *Anthemis arvensis* (280), *Hordeum murinum* (274), *Raphanus raphanistrum* (233), *Vicia sativa* (228), *Malva sylvestris* (197), *Sisymbrium officinale* (166) and *Echinochloa crus-galli* (165); while the most frequent neophytes were *Veronica persica* (212), *Erigeron canadensis* (138), *Amaranthus hybridus* (73), *Oxalis corniculata* (67), *Lepidium didymum* (57), *Oxalis latifolia* (51), *Paspalum distichum* (48), *Bidens frondosus* (35), *Cyperus eragrostis* (34) and *Symphyotrichum squamatum* (33).

By vegetation class, *Epilobietea angustifolii* had the highest proportion of natives in its species pool (82 %); *Sisymbrietea* had the highest proportion of putative archaeophytes (31 %); and *Bidentetea* had the highest proportion of neophytes (17 %). The alliances with the highest proportions of archaeophytes in their species pools were *Caucalidion lappulae* (41 %), *Chenopodion muralis* (35 %) and *Sisymbrion officinalis* (31 %); while the alliances with the highest proportion of neophytes were *Bidention tripartitae* (19 %), *Allion triquetri* (12 %) and *Spergulo arvensis-Erodion cicutariae* (10%).

@@

The most frequent life form in the species pool (n = 1,162 species) were hemicryptophytes (44 %), followed by therophytes (42 %), geophytes (9 %), chamaephytes (6 %), phanerophytes (6 %), bryophytes (1 %) and hydrophytes (1 %). The proportion of therophytes across classes and alliances (**Fig. 4**) agreed with the traditional description of the sintaxa. Thus, therophytes dominated the annual communities of trampled-soil vegetation (*Polygono-Poetea annuae*), crops weeds (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*) and annual ruderals (*Chenopodietea*, *Sisymbrietea*, *Bidentetea*); while perennial life forms (especially hemicryptophytes) dominated walls (*Cymbalario-Parietarietea diffusae*) and perennial ruderal vegetation (*Artemisietea vulgaris*, *Epilobietea angustifolii*). Geophytes were a relatively high proportion of the vegetation in three alliances: *Allion triquetri*, *Convolvulo arvensis-Agropyrion repentis* and *Senecionion fluviatilis*.

@@

To visualize the major patterns of variation in the ecological and disturbance preferences of the different communities (**Fig. 6**) we performed a PCA ordination of plot-level means (**Fig. 7**). The first two PCA axes explained 59 % of the variability. The major contributors to axis 1 (36 % variance explained) were light, nutrient and moisture requirements; it separated communities of open, dry and comparatively nutrient-poorer sites (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea*, *Artemisietea vulgaris*, *Chenopodietea* and *Digitario sanguinalis-Eragrostietea minoris*) from communities of shady, moist and nutrient-richer sites (classes *Cymbalario-Parietarietea diffusae*, *Bidentetea* and *Epilobietea angustifolii*). The major contributors to axis 2 (23% variance explained) were disturbance severity and frequency; this axis separated communities preferring low severity - low frequency disturbances (classes *Cymbalario-Parietarietea diffusae*, *Epilobietea angustifolii*, *Chenopodietea*, *Artemisietea vulgaris* and *Bidentetea*) from communities adapted to high severity - high frequency disturbances (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea* and *Digitario sanguinalis-Eragrostietea minoris*).

@@

We found a percentage of 22 % non-native species in the anthropogenic vegetation of the Iberian Atlantic ecoregion. This percentage is intermediate between that found in urban areas of temperate Europe and North America (30-50 %) [@RN5498; @RN5499; @RN5500; @RN5079] and Italy (12-26 %) [@RN5501]. A survey of anthropogenic vegetation in the north-west Balkans found 13 % non-native plants, with more neophytes than archaeophytes [@RN5429]. A similar study in the Czech Republic found 32 % archaeophytes and 7 % neophytes [@RN5435]. Taken together, these results suggest that the Iberian Atlantic anthropogenic vegetation shows an invasion status that is intermediate between southern and central-northern Europe, which is in agreement with the biogeographical position of our ecoregion. Previous studies also suggested that the invasion status was higher in early successional anthropogenic communities [@RN5423], and our results confirmed this as the proportion of native species tended to be higher in the perennial ruderal classes *Artemisietea vulgaris* and *Epilobietea angustifolii*. The classes with more alien species also tended to be those adapted to more frequent and more severe disturbances (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea*), also in agreement with previous research [@RN5435]. Of the alien species, archaeophytes were a higher proportion in the vegetation associated to arable habitats (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea*), in agreement with the ancient origin of archaeophytes during the development of agriculture [@RN5494; @RN5490]. On the other hand, neophytes tended to be in higher proportions in vegetation types associated to wet conditions (*Bidentetea*), as has been found in other European regions [@RN5497].

@@

The anthropogenic vegetation of the Iberian Atlantic ecoregion is home to 1,162 vascular plant taxa. This is approximately one third of the whole ecoregional species pool represented in the SIVIM database [@RN5374], or one fifth of the Iberian flora [@RN5503]. More than half of this synanthropic diversity (615 taxa) occurs in mesic to wet perennial ruderal vegetation, highlighting the potential of some anthropogenic habitats to act as source and supply of biodiversity in human-dominated landscapes [@RN5446; @RN5445]. These results suggest three main objectives for the sustainable management of human-made habitats: (1) focusing efforts on plant invasion management in anthropogenic plant communities of wet habitats; (2) conserving examples of archaeophyte-rich annual weed communities, especially those linked to traditional agricultural practices [@RN5502]; and (3) conserving and promoting mesic ruderal and fringe vegetation as a biodiversity refuge in urban and peri-urban areas.

>>>>

The total synanthropic species pool (n = 1,162) was composed of 78% natives (n = 908), 15% putative archaeophytes (n = 178) and 7% neophytes (n = 76). The most frequent putative archaeophytes were *Capsella bursa-pastoris* (390 occurrences), *Senecio vulgaris* (332), *Anisantha sterilis* (301), *Anthemis arvensis* (280), *Hordeum murinum* (274), *Raphanus raphanistrum* (233), *Vicia sativa* (228), *Malva sylvestris* (197), *Sisymbrium officinale* (166) and *Echinochloa crus-galli* (165); while the most frequent neophytes were *Veronica persica* (212), *Erigeron canadensis* (138), *Amaranthus hybridus* (73), *Oxalis corniculata* (67), *Lepidium didymum* (57), *Oxalis latifolia* (51), *Paspalum distichum* (48), *Bidens frondosus* (35), *Cyperus eragrostis* (34) and *Symphyotrichum squamatum* (33).

By vegetation class, *Epilobietea angustifolii* had the highest proportion of natives in its species pool (82%); *Sisymbrietea* had the highest proportion of putative archaeophytes (31%); and *Bidentetea* had the highest proportion of neophytes (17%). The alliances with the highest proportions of archaeophytes in their species pools were *Caucalidion lappulae* (41%), *Chenopodion muralis* (35%%) and *Sisymbrion officinalis* (31%); while the alliances with the highest proportion of neophytes were *Bidention tripartitae* (19%), *Allion triquetri* (12%) and *Spergulo arvensis-Erodion cicutariae* (10%).

@@

The most frequent life form in the species pool (n = 1,162 species) were hemicryptophytes (44%), followed by therophytes (42%), geophytes (9%), chamaephytes (6%), phanerophytes (6%), bryophytes (1%) and hydrophytes (1%). The proportion of therophytes across classes and alliances (**Fig. 4**) agreed with the traditional description of the sintaxa. Thus, therophytes dominated the annual communities of trampled-soil vegetation (*Polygono-Poetea annuae*), crops weeds (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*) and annual ruderals (*Chenopodietea*, *Sisymbrietea*, *Bidentetea*); while perennial life forms (especially hemicryptophytes) dominated walls (*Cymbalario-Parietarietea diffusae*) and perennial ruderal vegetation (*Artemisietea vulgaris*, *Epilobietea angustifolii*). Geophytes were a relatively high proportion of the vegetation in three alliances: *Allion triquetri*, *Convolvulo arvensis-Agropyrion repentis* and *Senecionion fluviatilis*.

@@

To visualize the major patterns of variation in the ecological and disturbance preferences of the different communities (**Fig. 6**) we performed a PCA ordination of plot-level means (**Fig. 7**). The first two PCA axes explained 59% of the variability. The major contributors to axis 1 (36% variance explained) were light, nutrient and moisture requirements; it separated communities of open, dry and comparatively nutrient-poorer sites (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea*, *Artemisietea vulgaris*, *Chenopodietea* and *Digitario sanguinalis-Eragrostietea minoris*) from communities of shady, moist and nutrient-richer sites (classes *Cymbalario-Parietarietea diffusae*, *Bidentetea* and *Epilobietea angustifolii*). The major contributors to axis 2 (23% variance explained) were disturbance severity and frequency; this axis separated communities preferring low severity - low frequency disturbances (classes *Cymbalario-Parietarietea diffusae*, *Epilobietea angustifolii*, *Chenopodietea*, *Artemisietea vulgaris* and *Bidentetea*) from communities adapted to high severity - high frequency disturbances (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea* and *Digitario sanguinalis-Eragrostietea minoris*).

@@

We found a percentage of 22% non-native species in the anthropogenic vegetation of the Iberian Atlantic ecoregion. This percentage is intermediate between that found in urban areas of temperate Europe and North America (30-50%) [@RN5498; @RN5499; @RN5500; @RN5079] and Italy (12-26%) [@RN5501]. A survey of anthropogenic vegetation in the north-west Balkans found 13% non-native plants, with more neophytes than archaeophytes [@RN5429]. A similar study in the Czech Republic found 32% archaeophytes and 7% neophytes [@RN5435]. Taken together, these results suggest that the Iberian Atlantic anthropogenic vegetation shows an invasion status that is intermediate between southern and central-northern Europe, which is in agreement with the biogeographical position of our ecoregion. Previous studies also suggested that the invasion status was higher in early successional anthropogenic communities [@RN5423], and our results confirmed this as the proportion of native species tended to be higher in the perennial ruderal classes *Artemisietea vulgaris* and *Epilobietea angustifolii*. The classes with more alien species also tended to be those adapted to more frequent and more severe disturbances (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea*), also in agreement with previous research [@RN5435]. Of the alien species, archaeophytes were a higher proportion in the vegetation associated to arable habitats (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea*), in agreement with the ancient origin of archaeophytes during the development of agriculture [@RN5494; @RN5490]. On the other hand, neophytes tended to be in higher proportions in vegetation types associated to wet conditions (*Bidentetea*), as has been found in other European regions [@RN5497].

@@

The anthropogenic vegetation of the Iberian Atlantic ecoregion is home to 1,162 plant taxa. This is approximately one third of the whole ecoregional species pool represented in the SIVIM database [@RN5374], or one fifth of the Iberian flora [@RN5503]. More than half of this synanthropic diversity (615 taxa) occurs in mesic to wet perennial ruderal vegetation, highlighting the potential of some anthropogenic habitats to act as source and supply of biodiversity in human-dominated landscapes [@RN5446; @RN5445]. These results suggest three main objectives for the sustainable management of human-made habitats: (1) focusing efforts on plant invasion management in anthropogenic plant communities of wet habitats; (2) conserving examples of archaeophyte-rich annual weed communities, especially those linked to traditional agricultural practices [@RN5502]; and (3) conserving and promoting mesic ruderal and fringe vegetation as a biodiversity refuge in urban and peri-urban areas.
